---
title: "Exploring Hillary et al. and the estimate of White Shark Abundance, etc."
output: html_notebook
bibliography: references.bib
---

Here we explore the models used in @hillary2018genetic.  Looking through the paper, I didn't
find any direct mention of a public repository of data that was used in the paper,
nor any repository of code for their analyses. So, we are just going to simulate
some data here from their model (first just from their inference model, and maybe later
we will do so with a simple individual-based-model) so that we have something to work with.
One nice thing about simulating from their simulation model is that doing so offers a good
way to gain a better understanding their model.

## Their inferential model

They start with a demographic model of exponential growth:
$$
N_t^A = N_\mathrm{init}e^{\lambda t}
$$
where $N_t^A$ is the number of adults at time $t$ and $N_\mathrm{init}$ is the initial number
of adults (at time $t = 0$). $\lambda$ is an exponential growth rate.  

Here, we define a function to calculate $N_t^A$:
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
N_adults <- function(N0, t, lambda) {
  N0 * exp(lambda * t)
}
```

Let's see what that would look like over 60 years, starting from a few differenty $N_\mathrm{init}$
values and growing/shrinking at a range of values.
```{r}
A_traj <- expand_grid(
  Ninit = c(600, 800, 1000, 1200, 1500, 2000), 
  lambda = seq(-0.03, 0.03, by = 0.01)
)%>%
  mutate(
    num_adults = map2(Ninit, lambda, function(x, y) N_adults(x, 0:60, y)),
  ) %>%
  unnest(num_adults) %>%
  group_by(Ninit, lambda) %>%
  mutate(year = 0:60)

ggplot(A_traj, aes(x = year, y = num_adults, colour = factor(lambda))) +
  geom_line() +
  facet_wrap(~ Ninit)

```
Now that we have all those different trajectories, let's nest them
into separate tibbles so we can operate easily on one at a time with
different functions.

```{r}
A_traj_nest <- A_traj %>%
  group_by(Ninit, lambda) %>%
  nest()
```

And now we are going to simulate some half-sib pair data from each of these
models, pretending that we have sampled 10 sharks between ages 3 and 8 each
year from years 20 to 40.  We will assume the sample is of the same size 
and ages for each scenario.

So, first, we sample individuals from each different scenario.
```{r}
set.seed(5)
samples <- tibble(
  samp_year = rep(20:40, each=10)
) %>%
  mutate(
    age = sample(3:8, size = n(), replace = TRUE, prob = c(8:3)),
    born_year = samp_year - age
  )
```

So, that is 210 samples.  Now, let's join those on with the population trajectories, and
then nest them all into separate tibbles of samples and the population sizes during the
year when they were born
```{r}
multi_samples <- samples %>%
  left_join(A_traj, by = c("born_year" = "year")) %>%
  group_by(Ninit, lambda) %>%
  nest() %>%
  rename(samples = data)
```

Now, we want to compute the probability that each of the possible pairs of those samples
are HSPs from the number of adults when the younger member of the pair was born.
Let's first form all the pairs of them:
```{r}
# make a function to create the pairs
make_pairs <- function(samples) {
  n <- nrow(samples)
  s1 <- samples[rep(1:n, each = n),]
  s2 <- samples[rep(1:n, times = n),]
  names(s2) <- paste0(names(s2), ".old")
  
  pairs <- bind_cols(s1, s2) %>%
    filter(born_year.old < born_year)
  pairs
}

# then apply that:
multi_pairs <- multi_samples %>%
  mutate(pairs = map(samples, make_pairs))
```

Now, for each of these pairs we want to compute the half-sibling probability.
Doing so requires the adult survival probability.  Tell you what---let's
compute these for a range of different adult survival probabilities.

First, we need to write a function for the HSP probabilities.
```{r}
HSP_probs <- function(num_adults, phiA, born_year, born_year.old) {
  (4/num_adults) * (phiA ^ (born_year - born_year.old))
} 
```
Then we can go ahead and apply those:
```{r}
multi_pairs2 <- expand_grid(phiA = c(0.85, 0.90, 0.94, 0.96), multi_pairs) %>%
  mutate(pairs = map2(pairs, phiA, function(x, y) x %>% mutate(HSP_prob = HSP_probs(num_adults, y, born_year, born_year.old))))
  
```

OK, that is a big mongo-sized tibble of lots of HSP probs under different scenarios.
Let's plot a few cases. Try this:
```{r, fig.height=4, fig.width=4, out.width='100%'}
multi_pairs2 %>% filter(
  near(phiA, 0.94) | near(phiA, 0.9) | near(phiA, 0.85), 
  Ninit == 800, 
  near(lambda, 0.01) | near(lambda, 0) | near(lambda, -0.01)
) %>% unnest(pairs) %>%
  ggplot(
  ., 
  aes(
    x = born_year, 
    y = HSP_prob, 
    fill = factor(born_year - born_year.old), 
  )
) +
  geom_point(shape = 21) +
  facet_grid(lambda ~ phiA) +
  theme(legend.position = "none")
```

Yep, that totally makes sense.

So, now we want to basically use the HSP prob from each condition and use it to simulate
whether each pair is an HSP or not, and once we have done that we will count up how many
HSPs (and not) there are from each scenario.

```{r}
mp3 <- multi_pairs2 %>%
  mutate(pair_counts = map(
    pairs, 
    function(x) {
      x %>% 
        mutate(
          isHSP = rbernoulli(n(), p = HSP_prob),
          born_year_diff = born_year - born_year.old
        ) %>%
        count(samp_year, born_year_diff, HSP_prob, isHSP)
    }))
```
# Citations

