---
title: "CKMR in Age-Structured Populations: a Pictorial Introduction"
output: pdf_document
author: "Eric C. Anderson"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

\newcommand{\Fec}{\mathrm{Fec}}
\newcommand{\bE}{\boldsymbol{E}}
\newcommand{\bN}{\boldsymbol{N}}
\newcommand{\bS}{\boldsymbol{S}}



When I am working on CKMR in age-structured populations it always seems
that I end up having to re-derive a lot
of the kinship probabilities, and I always wish that I had a nice tidy expression for
kinship probabilities, like Paul had in his presentation at the TWS workshop.
Or, I might be working on some sort of new weird case,
in which case I actually really do have to derive the probabilities.  In both cases,
I always wish that I had better picture in my head of the quantities involved.

This document provides a derivation of some common (and eventually somewhat more obscure)
kinship probabilities in a pictorial format.  I have found it helpful for me to have
a picture to look at things, and it often makes it much easier for me to verify things
like limits of summation, etc.

The code for making these little pictures, and such, is available in my R package
'ckviz' available at [https://github.com/eriqande/ckviz](https://github.com/eriqande/ckviz).



# Vital rates, age-specific abundance and ERRO

We will consider a situation with organisms having a maximum age of $a_\mathrm{max}$ and we
will consider times from 1 to $T$.  The most refined target of inference
using CKMR in this scenario would be
$N_{s,t,a}$, the number of individuals of sex $s$ at time $t$ of age $a$.  

Let's take a concrete example with $T=15$ and $A=a_\mathrm{max}=8$.  And let's also, for the
sake of illustration, make up some fairly simple life-history parameters, given in the following
table.  $S_a$ (written `S_a` in the table) is the probability of
surviving from age $a-1$ to $a$, and, for simplicity
here, it is assumed
to be the same for males and females.  $M_m$ (`M_m`) is the fraction of males of each age that
are mature, while $M_f$ (`M_f`) is the fraction of females that are mature.  Meanwhile,
`Fec_f` is the fecundity of females, and `Fec_m` is the expected reproductive
potential of each male of a given age. Note that these last two are not given on
a relative scale, but they could have been.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ckviz)  # get it from github at eriqande/ckviz

vitals <- example_vitals()
knitr::kable(vitals[1:8,], digits = 2)
```

Then, if that system had a mild amount of variation in the number of
age-0 offspring produced, as well as a bit of variation in the realized
survival rates each year for each age group, we might be able to represent
the number of individuals of sex $s$ of each age $a$ in each year
$t$, pictorially, with something like this:

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=2.2}


set.seed(5)
Es <- create_Es_from_vitals(vitals = vitals, T = 15)


plot_ckmr_pictorial(E = Es, cell_fill_var = "Nt", cell_str_var = "N")
  
```

It is worth noting that the recruitment variability (number of 0-year-olds that make it
to be 1-year-olds) is evident in the streaks of lighter and darker colors moving
diagonally to the right up the plot.  

Those $N$'s are, of course, unknown.  We will be trying to estimate something like those $N$'s, but
we won't estimate each different $N_{s,t,a}$.  Rather we will have a model (perhaps like a sort of
Leslie Matrix, or maybe something parameterized in combination with an exterally estimated
age composition---or maybe a combination of both) by which we can estimate all
those $N_{s,t,a}$, based on a much smaller number of parameters.  

When it comes time to compute kin-probabilities in this model, some related quantities
that we can use are the year-specific age compositions.  We use $R_{s,t,a}$
to denote the relative abundance of $a$-year-olds of sex $s$ in year $t$.  Most simply
we can think of these as proportions:
$$
R_{s,t,a} = \frac{N_{s,t,a}}{\sum_{a'} N_{s,t,a'}}
$$

Based on the simulated $N$ values above, here is what those $R$'s look like:

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=2.2}

plot_ckmr_pictorial(E = Es, cell_fill_var = "R", cell_str_var = "Rlab")

```

The reason that these $R$ values are relevant for calculating the kin
probabilities is that every direct kin-probability calculation that you will do is going
to be conditioned on the birth years of the first-born and the second-born individuals
in the pair.  (Sometimes there might be uncertainty in those years, but to deal with that
we will be taking a sum over the unobserved birth years, so you will still be
conditioning on those birth years!)

Conditioning upon a certain birth year, the
chance that an individual was born to an individual of a given age depends on the
_expected relative reproductive output_, or ERRO, of all the individuals of that age, and that
ERRO can be equivalently computed using the $R$'s or the $N$'s.  It can be useful to formulate
the ERRO in terms of the $R$'s rather than the $N$'s because you might have age
composition estimates from some other sources and those could be used to inform
the $R$'s in the model, etc.

We will use $E_{s,t,a}$ to denote the ERRO of individuals of sex $s$ and age $a$
at time $t$.  Note that these quantities are fractions of expected reproductive
output relative to all the reproducing individuals of sex $s$ at time $t$.  They
are calculated as:
$$
E_{s,t,a} = \frac{R_{s,t,a} M_{s,a} \Fec_{s,a}}{\sum_{a'} R_{s,t,a'} M_{s,a'} \Fec_{s,a'}},
$$
where $M_{s,a}$ is the probability that an $a$-year-old of sex $s$ is mature and
$\Fec_{s,a}$ is the expected (relative, if desired) reproductive output of a
mature $a$-year-old of sex $s$.
For females this is often taken to be the fecundity.

So, here is what the ERRO looks like for females in our simple example:

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=2.2}
# let us make a note of the maximum and minimum ERROs here, because
# we will want the fill color scale to remain constant acccording to those
max_erro <- max(c(Es$Emt, Es$Eft), na.rm = TRUE)
min_erro <- min(c(Es$Emt, Es$Eft), na.rm = TRUE)


plot_ckmr_pictorial(E = Es, cell_fill_var = "Eft", cell_str_var = "Ef_str", min_fill = min_erro, max_fill = max_erro)

```


and here it is for males:


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=2.2}

plot_ckmr_pictorial(E = Es, cell_fill_var = "Emt", cell_str_var = "Em_str", min_fill = min_erro, max_fill = max_erro)

```

White cells indicate no reproductive output.  Those are the ages where females
are not yet reproductively mature.  Some fraction of the males mature at age
1 and 2, so there are no white cell amongst the males.


# Calculating probabilities that pairs are kin

Now that we have these basic pictures set, we can start using them to understand
the process whereby we calculate the probability that two samples---call them
$K_1$ and $K_2$---that were
born in years $b_1$ and $b_2$, respectively, with $b_1 < b_2$,
are related to one another.  

It is important to note here that we are usually going to know the birth years
of the members of the pair because:

1. We know when we sampled the individual.
2. We know its age when we sampled it.

Thus, it is really important to be able to accurately age the individuals in your
CKMR sample.  Sometimes, of course, there might be uncertainy around the age of
the individual.  In that case, one has to sum over the different possible ages of the
members of a pair, weighting by their age probabilities.  That is merely an extra
layer that sits above what we are doing here.  For each term in the sum over ages,
there is a probability that is calculated given that the age of the individual is
known, and those probabilities are what we are calculating here.

## Half-sibling probabilities

We will start with
the case of half-sibling relationships.  To make an initial picture of this
we will take a concrete example, with
the first-born ($K_1$) being born in year $b_1 = 4$ and the second born ($K_2$)
being born in year $b_2=6$,  and we will
think about the probability that the pair are maternal half-sibs.  

One important thing to note is this: we are conditioning on
$K_1$ and $K_2$ having survived until whatever
time they were sampled, so we don't have to include any terms for the samples
themselves surviving any number of years.

When we calculate these sorts of probabilities we are always conditioning
upon the birth years of the pair, which means that we focus on a subset of the
adult population that the offspring could have come from.  In this case, that
subset is, "all pairs of samples in which the first-born was born at time
4 and the second-born was born at time 6."  Thus, we can represent the
females in the population that could have given rise to this subset of
pairs like this:

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=2.2}

hsp_mask1 <- expand_grid(t = c(4, 6), a = 1:8)
plot_ckmr_pictorial(
  E = Es, cell_fill_var = "Eft", 
  cell_str_var = "Ef_str", 
  min_fill = min_erro, 
  max_fill = max_erro, 
  cell_mask = hsp_mask1
)


```

The colored cells above are showing us all the components of the female population that
could be mothers of a pair
in which the first-born was born in year 4 (the first column of colored cells) and the
second-born was born in year 6 (the second column of colored cells).  Remember,
that represents our "universe"---with probability one, a pair with $b_1=4$ and $b_2=6$
came from mothers in those colored segments of the population.

In order to calculate the probability that a pair consists of maternal half siblings we start
carving out fractions of that probability of one by invoking all the events that must occur
in order for the pair to be maternal half-siblings. Those events can be
summarized like this:

1. The mother that produced $K_2$ at time $b_2$ must be $b_2 - b_1$ years older
than the mother that produced $K_1$.  (Clearly, if $K_2$'s mother is not $b_2-b_1$
years older than $K_1$'s mother, then there is no chance that those two mothers
are the same individual).
2. The mother that produced $K_1$ must have been present at the time of $K_2$'s
birth.  For this to happen, $K_1$'s mother must have survived the intervening years.
3. Given the first two events happened, finally, it must be the case that
the mother of $K_1$ is also the mother of $K_2$.  Since all of the
$N_{f,t,a}$ females are assumed to have the same ERRO, the probability
that the mother of $K_1$ is the mother of $K_2$ is just the expected fraction
of all the $N_{f,t,a}$ mothers in a cell that are from a single one of those
mothers, and that is simply $1/N_{f,t,a}$.  

In order to handle all these probabilities elegantly and come up with
a single number that is the maternal half-sibling probability, we typically
have to sum these sorts of probabilities over all the possible ages, $a^*$,
of the mother of the first-born.  Therefore, we are going to present graphical
images for each of the three above events/probabilities for the particular case
in which we are conditioning upon the age of the mother of $K_1$ being $a^* = 5$.

Thus, we can depict event 1, with $a^* = 5$ like this:

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=2.2}

hsp_mask2 <- tibble(t = c(4, 6), a = c(5, 7))
plot_ckmr_pictorial(
  E = Es, cell_fill_var = "Eft", 
  cell_str_var = "Ef_str", 
  min_fill = min_erro, 
  max_fill = max_erro, 
  cell_mask = hsp_mask2
)



```

That is merely showing the event in which first-born from time $b_1 = 4$ came from a mother of age $a^* = 5$
and the second born from time $b_2 = 6$ came from a mother of age 7 at time $b_2$. The important part to
see here is that the probability of finding a pair with birth years $b_1$ and $b_2$ that came from
mothers of those ages is, simply:
$$
E_{f,4,5}E_{f,6,7}.
$$

The second event that has to happen is that the mother of the first-born must survive
at least until the time of the second-born's birth.  We are going to depict that with
a line, whose width denotes the probability that this mother has survived through these
different time steps:

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=2.2}


plot_ckmr_pictorial(
  E = Es, cell_fill_var = "Eft", 
  cell_str_var = "Ef_str", 
  min_fill = min_erro, 
  max_fill = max_erro, 
  cell_mask = hsp_mask2, 
  survival_endpoints = tibble(t1 = 4, a1 = 5, t2 = 6, a2 = 7)
)



```

It is a little hard to see it, but that line gets skinnier as it moves into new
cells going up and to the right.  The width in the cell with $K_1$'s mother
($t = 4$, $a = 5$) denotes a probability of 1.0 (she existed at that time) and in
every time increment going forward, the linewidth represents the cumulative
probability of surviving to that cell, on a linear scale.  In other words, the
probability of surviving to the cell
at $t=5$ and $a=6$ is $S_6 = 0.95$,
and the probability of surviving all the way to age 7 is $S_6 * S_7 = 0.95 * 0.95 = 0.9025$.
The line widths reflect those probabilities.



Just for fun, because the changes in line width will be easier to see there,
let's make a picture for male parents of a pair with $b_1 = 5$ and
$b_2 = 11$.
The two possible ages for a father of the first born that then survived
for $11 - 5 = 6$ more years are
then 1 and 2, and we get this:

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=2.2}

cm3 <- tibble(t = c(5, 5), a = c(1, 2)) %>%
  bind_rows(tibble(t = c(11, 11),  a = c(7, 8)))
segs <- tibble(
  t1 = c(5, 5), a1 = c(1, 2)
) %>%
  mutate(t2 = t1 + 6, a2 = a1 + 6)

plot_ckmr_pictorial(
  E = Es, 
  cell_fill_var = "Emt", 
  cell_str_var = "Em_str", 
  min_fill = min_erro, 
  max_fill = max_erro, 
  cell_mask = cm3, 
  survival_endpoints = segs
)



```

Yes, that is nice to look at. Enough years are passing so that the differences
in probability of survival from the beginning to the end are more evident. Also,
the lower line is much skinnier at the end than the upper line because survival
from age 1 to age 7 is lower than it is from age 2 to 8.  

Anyways, let us return to our simple example with $a^* = 5$.  We are almost
there.  Up to this point, we have depicted events 1 and 2, and they have a
joint probability of:
$$
E_{f,4,5}E_{f,6,7} S_6S_7.
$$
The last event that must occur is for the mother of $K_2$ to be the same individual
as the mother of $K_1$.  Because each of the mothers of age 7 at time 6 has the same
ERRO, they are each pumping out, on average, the same number of offspring, and the
probability that $K_2$'s mother is a specific one of those females ($K_1$'s mother,
to be exact!) is just proportional to the average number of offspring being pumped out by any
single individual, as a fraction of all the offspring getting produced by females
of age 7 at time 6.  Since there are $N_{f,6,7}$ females pumping out offspring, the
probability of sampling an offspring from a specific female is simply $1/N_{f,6,7}$.
We will represent this with a small circle which symbolically (not in terms of
total area!) represents the subset of offspring that came from
the mother of $K_1$.  Hence our picture representing all three events (and their
probabilities) now looks like:

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=2.2}

plot_ckmr_pictorial(
  E = Es, cell_fill_var = "Eft", 
  cell_str_var = "Ef_str", 
  min_fill = min_erro, 
  max_fill = max_erro, 
  cell_mask = hsp_mask2, 
  survival_endpoints = tibble(t1 = 4, a1 = 5, t2 = 6, a2 = 7),
  circ_cells = tibble(t = 6, a = 7)
)

```

And the probability that represents is:
$$
\frac{E_{f,4,5}E_{f,6,7}}{N_{f,6,7}} S_6S_7.
$$

Of course, the calculation depicted above is just getting us the
joint probability that $K_1$ and $K_2$ are maternal half-siblings
_and_ their shared mother was of age $a^*=5$ in the year $b_1$.  To get
the actual kin probability we have to sum those joint probabilities over all
possible values of $a^*$.  The picture of all those probabilities looks like:

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=2.2}

starts = tibble(t = rep(4, 4), a = 3:6)
cm_hsp_full <- bind_rows(
  starts,
  starts %>% mutate(t = t + 2, a = a + 2)
)
spts <- starts %>%
  rename(t1 = t, a1 = a) %>%
  mutate(t2 = t1 + 2, a2 = a1 + 2)

plot_ckmr_pictorial(
  E = Es, cell_fill_var = "Eft", 
  cell_str_var = "Ef_str", 
  min_fill = min_erro, 
  max_fill = max_erro, 
  cell_mask = cm_hsp_full, 
  survival_endpoints = spts,
  circ_cells = spts %>% mutate(t = t2, a = a2) %>% select(t, a)
)

```

This picture and development should now make clear how we can write the maternal half-sibling kin probability,
more generally. Let $a_\mathrm{min}$ be the first age at which a female has non-zero probability of producing offspring (3, in
our example) and $a_\mathrm{max}$ be the maximum age at which a female will produce offspring (this is our maximum age
of 8, in the example), and let $d = b_2 - b_1$, be the number of years between the births. Then, we can write:
$$
P(K_1~\mbox{and}~K_2~\mbox{are}~\mathrm{MHS}|b_1,b_2,\bE, \bN, \bS) = \sum_{a^*=a_\mathrm{min}}^{a_\mathrm{max} - d}\biggl[
\frac{E_{f,b_1,a^*}E_{f,b_2,a^*+d}}{N_{f,b_2,a^*+d}}
\prod_{a=a^*+1}^{a^* + d} S_a
\biggr]
$$

Voila!


## Parent-offspring probabilities

Next we will derive these...

## Other relationships

Although half-sibling and parent-offspring pairs are the ones more readily
used in a CKMR experiment, there are other relationships that are important---most
often because these other relationships can be mistaken for the target relationship
of half-sibling.  Being able to calculate the probabilities of their occurrence
can help to account for their presence in the data set.  This should be fun.

### Grandparent-grandchild

Gotta do this still.

### Half-thiatic (i.e.  half aunt-niece)

Gotta do this one, also. Even though these are not target kin pairs, typically,
it would be nice to know that prbability of their occurrence---especially since
that would be a critical step to actually dealing with their presence in a probabilitic
fashion, rather than just setting the false negative rate for half-siblings so high
that you don't expect to see any half-thiatics in there.

Note that this one ends up needing information
about increase in kinship probabilities for individuals that are are born
in the same year.  We will discuss that when we talk about full siblings.

### Full siblings

As promised, we will discuss overdispersed reproductive success in this context.
Or perhaps first we will talk about within-cohort and cross-cohort half-sibling
comparisons, and ways to parameterize the corresponding probabilities by a
scaled variance in reproductive success parameter.  

# Modeling the underlying quantities

From the foregoing, we see that much of the art that goes into combining CKMR with
additional fisheries data involves using those additional data to inform
the variables that appear in the expresssions and pictures above.  Namely the:

- year-specific age compositions (these inform the $R$'s and hence the $N$'s)
- a recruitment index (these will inform the $R$'s and $N$'s, too)
- estimates of fishing mortality (these will inform the $S$'s)
- etc.

This is the stuff that Claudia has done so beautifully.  

